---
title: "Transmission Type vs Miles per Gallon Analysis"
author: "Marcus Vinicius Monteiro de Souza"
date: "28 de junho de 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this project we'll analyze the [Motor Trend Car Road Tests](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) data set to answer the following questions:

1. Is an automatic or manual transmission better for MPG?
2. How can we quantify the MPG difference between automatic and manual transmissions?

## Preparations

In this section, we'll install and load the required [R](https://www.r-project.org/) [packages](https://cran.r-project.org/web/packages/):

```{r install_and_load_required_packages}
if (!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(GGally, lsr, plyr, tidyverse)
```

and load and preprocess the data for the analysis:

```{r load_and_preprocess_mtcars_data}
data("mtcars")

mtcars <- mtcars %>%
  mutate(cyl = as.factor(cyl), vs = as.factor(vs), 
         am = as.factor(am), am = revalue(am, c('0' = 'automatic', '1' = 'manual')),
         gear = as.factor(gear), carb = as.factor(carb))
```

## Exploring the data

The `mtcars` data has 32 observations of the following 11 variables:

| Variable Name | Description                                              |
| --------------|:---------------------------------------------------------|
| mpg           | Miles/(US) gallon                                        |
| cyl           | Number of cylinders                                      |
| disp          | Displacement (cu.in.)                                    |
| hp            | Gross horsepower                                         |
| drat          | Rear axle ratio                                          |
| wt            | Weight (1000 lbs)                                        |
| qsec          | 1/4 mile time                                            |
| vs            | V/S                                                      |
| am            | Transmission (0 = automatic, 1 = manual)                 |
| gear          | Number of forward gears                                  |
| carb          | Number of carburetors                                    |

The observations look like this:

```{r head_mtcars}
head(mtcars)
```

Let's see a raw data [box plot](http://www.physics.csbsju.edu/stats/box2.html) of `mpg` by transmission type:

```{r mpg_by_am_boxplot}
 mtcars %>%
  ggplot(aes(am, mpg)) + 
  geom_boxplot(aes(fill = am)) +
  labs(title = 'Miles per Gallon by Transmission Type', x = 'Transmission Type', 
       y = 'Miles per Gallon', fill = 'transmission type')
```

Apparently, manual transmission cars do more miles per gallon than automatic transmission ones... but is it really? Let's analyze it.

## Is an automatic or manual transmission better for MPG?

To answer this question, we'll statistically compare the means of `mpg`s of both automatic and manual transmission populations. First, we'll check if `mpg` can be considered as normally distributed:

```{r mpg_qqplot}
qqnorm(mtcars$mpg, main = 'Mpg Normal Q-Q Plot')
qqline(mtcars$mpg, col = 2)
```

`mpg`'s distribution appears normal. Now, because each observation was taken independently, let's do a [unpaired, or independent samples test](https://en.wikipedia.org/wiki/Student%27s_t-test#Unpaired_and_paired_two-sample_t-tests):

```{r mpg_independent_samples_t_test}
independentSamplesTTest(mpg ~ am, mtcars)
```

The conclusion is: the mean `mpg` of automatic transmission cars was *17.147 (std dev = 3.834)*, whereas the mean `mpg` of manual transmission ones was *24.392 (std dev = 6.167)*. A Welch's independent samples t-tests showed that this *7.245* miles per gallon difference between the populations is was **significant** *(t(18.332) = -3.767, p = .001, CI95 = [-11.28, -3.21], d = 1.411)*, rejecting the null hypothesis that population means equal for both groups and suggesting that **manual transmission cars are better for `mpg`**.

## How can we quantify the MPG difference between automatic and manual transmissions?

To answer this question we'll [fit a multiple regression model](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html) for `mpg`, [check the residuals](http://blog.minitab.com/blog/adventures-in-statistics-2/why-you-need-to-check-your-residual-plots-for-regression-analysis) and then draw our conclusion.

### Model Selection

```{r mpg_model_fit}
mpg.model <- lm(mpg ~ ., mtcars)
# Select a formula-based model by AIC
mpg.model <- step(mpg.model, trace = FALSE)
summary(mpg.model)
```

### Checking the residuals

```{r mpg_model_checking_the_residuals}
ggnostic(mpg.model, aes(color = am))
```

We can see that the residuals for `am` are within the lines, appearing to be normally distributed. We can now draw our conclusion.

### Conclusion

Our final model has Adjusted R-squared value of 0.8401, meaning that it explains 84% of the variation in `mpg`. According to it, **manual transmission cars can run 1.80921 miles per gallon more than automatic transmission ones**.
